---
title: "Cleaning data"
author: "Susi"
date: "2026-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load / install required packages

load_pkg <- function(pkg){
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Load required packages
load_pkg("readxl")
load_pkg("dplyr")
load_pkg("ggplot2")
load_pkg("tidyr")
load_pkg("janitor")
load_pkg("here")
load_pkg("ggpubr")
load_pkg("knitr")


```

## Precleaning in EXCEL

- adjustment of . vs , for numbers, captial letters vs small.
- fixed copy/paste error for SZajitschek data
- Libor's data: 30f10 had N/As for "latency" - I copied the values from the full 30 conditions, as that should be identical
- Helen's data: newly merged what was in google sheet and in separately submitted file; separation / re-coding of conditions (10|1, 30|1, 30f10). PROBLEM: when merging, it appears that original 10 and 30 conditions were analysed again, however with slightly different results. As this isn't mere duplication I have left those duplicated values in there (may give some interesting insight on differences when re-coring same videos?), but called these with less information 10|b and 30|b respectively (these conditions will be ignored in the analyses in comparison across species, but could be called separately specifically to compare outcomes for the double-scorong!)

## Import data

```{r}

data <- read_excel(here("raw data/combined_10March_2026.xlsx"))
str(data)

```
## Fixing column names

```{r}

data <- data %>% clean_names()
names(data)

```

```{r}

# Structure 
str(data)

# Converting inconsistent "N/A" and "-" strings to NA
data <- data %>% mutate(across(everything(), ~ ifelse(. %in% c("N/A", "-", ""), NA, .)))

# checking for duplicate rows
sum(duplicated(data)) # none


# as per structure, many variable are character but should be either numeric or factors
# converting to factors:
  
factor_vars <- c("study_species", "individual_id", "sex", "age", "origin", "strain", "condition", "trial_day", "background_colour", "housing_s", "housing_social")

data <- data %>%
  mutate(across(all_of(factor_vars), as.factor))

# converting to numeric:
  
number_vars <- c( "latency_swim_freely", "timespent_contact" ,"timespent_active", "nb_dart_startle", "distance_no","latency_approach_no", "latency_touch_no", "luminosity", "photoperiod", "salinity")

data <- data %>%
  mutate(across(all_of(number_vars), as.numeric))


summary(data)



```

```{r}

data %>% filter(trial_day == 1, condition == 10) %>%
  count(study_species) %>%
  ggplot(aes(x = study_species, y = n, fill= study_species)) +
  geom_col() +
  labs(title="Sample sizes by Species", x="Species", y="Count")

```
This time restricted to Day one and one condition, as we have duplicated rows. This shows numbers of individuals per species.

## assessing condions
```{r}
unique(data$condition)

```
This is weird, am assuming that the "I" is in there accidentally, instead of "|". Changing this requires reformatting to character again, changin the letter and then making it factor again.

```{r}

data$condition <- as.character(data$condition)
data$condition <- gsub("I", "|", data$condition, fixed = TRUE)
data$condition <- factor(data$condition)

```

ok, better, now check how many we have in each condition
```{r}
table(data$condition)
```
Who is missing?
```{r}

df <- data %>%
  mutate(unit_id = paste(researcher_name, study_species, individual_id, trial_day, sep = " | "))


keys_10    <- df %>% filter(condition == "10")    %>% distinct(unit_id, researcher_name, study_species)
keys_10_1  <- df %>% filter(condition == "10|1")  %>% distinct(unit_id)

missing_10_1 <- keys_10 %>%
  filter(!unit_id %in% keys_10_1$unit_id)
# great. issues with IDs when copying over!! in SZ data


keys_30     <- df %>% filter(condition == "30")    %>% distinct(unit_id, researcher_name, study_species)
keys_30_1   <- df %>% filter(condition == "30|1")  %>% distinct(unit_id)
keys_30f10  <- df %>% filter(condition == "30f10") %>% distinct(unit_id)

missing_30_1 <- keys_30 %>%
  filter(!unit_id %in% keys_30_1$unit_id)

missing_30f10 <- keys_30 %>%
  filter(!unit_id %in% keys_30f10$unit_id)



summary_missing <- full_join(
  missing_10_1  %>% count(researcher_name, study_species, name = "missing_10_1"),
  missing_30_1  %>% count(researcher_name, study_species, name = "missing_30_1"),
  by = c("researcher_name","study_species")
) %>%
  full_join(
    missing_30f10 %>% count(researcher_name, study_species, name = "missing_30f10"),
    by = c("researcher_name","study_species")
  ) %>%
  mutate(across(starts_with("missing_"), ~replace_na(.x, 0L))) %>%
  arrange(desc(missing_30f10 + missing_30_1 + missing_10_1))

summary_missing



```

what do we have?
```{r}

counts_long <- as.data.frame(with(data, table(researcher_name, study_species, condition)))
counts_long <- subset(counts_long, Freq > 0)
counts_long <- counts_long[order(counts_long$researcher_name), ]
kable(counts_long)

```

Quick recap: Joachim: no subsets prvided yet. Only 10 / 30

Maria: missing full 30, last mins for both 10 and 30 (but has provided 10 and first 10 for 30)

Melissa: Patchy data for the subsets. Video problems (?)

Helen: duplicated data, hope to have taken care of that.

```{r}

with(data, table(condition, trial_day))
     
   
```



let's have a quick look at conditions (last min 10 vs 30)

## Last Minute
```{r}
sub_last <- subset(data, condition %in% c("10|1", "30|1"))

ggplot(sub_last, aes(x = trial_day, y = timespent_active, colour = condition, fill = condition)) +
  geom_violin(
    position = position_dodge(width = 0.8),
    alpha = 0.25,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    alpha = 0.35
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    alpha = 0.45,
    size = 1.2
  ) +
  labs(x = "Trial day", y = "Time spent active (s)", colour = "Condition", fill = "Condition") +
  theme_minimal(base_size = 13)
```

### split by species: LAST minute only
```{r}
ggplot(sub_last, aes(x = trial_day, y = timespent_active, colour = condition)) +
  geom_line(
    aes(group = interaction(individual_id, condition)),
    alpha = 0.12,
    position = position_dodge(width = 0.8)
  ) +
  geom_violin(
    aes(fill = condition),
    position = position_dodge(width = 0.8),
    alpha = 0.2,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    alpha = 0.3
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    alpha = 0.4,
    size = 1.1
  ) +
  facet_wrap(~study_species) +
  labs(x = "Trial day", y = "Time spent active (s)", colour = "Condition", fill = "Condition", title = "LAST minute only - active") +
  theme_minimal(base_size = 13)

```


## 10 mins

```{r}
sub10 <- subset(data, condition %in% c("10", "30f10"))

ggplot(sub10, aes(x = trial_day, y = timespent_active, colour = condition)) +
  geom_line(
    aes(group = interaction(individual_id, condition)),
    alpha = 0.12,
    position = position_dodge(width = 0.8)
  ) +
  geom_violin(
    aes(fill = condition),
    position = position_dodge(width = 0.8),
    alpha = 0.2,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    alpha = 0.3
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    alpha = 0.4,
    size = 1.1
  ) +
  facet_wrap(~study_species) +
  labs(x = "Trial day", y = "Time spent active (s)", colour = "Condition", fill = "Condition", title = "First 10min acitivity") +
  theme_minimal(base_size = 13)

```
Neolamprologus = Joachim, DATA NOT COMPLETE YET (i.e. only full 10 / full 30 min available)


### Different variables?
LATENCY
```{r}

sub_lat <- subset(data, condition %in% c("10", "30f10"))

ggplot(sub_lat, aes(x = trial_day, y = latency_swim_freely, colour = condition)) +
  geom_line(
    aes(group = interaction(individual_id, condition)),
    alpha = 0.12,
    position = position_dodge(width = 0.8)
  ) +
  geom_violin(
    aes(fill = condition),
    position = position_dodge(width = 0.8),
    alpha = 0.2,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    alpha = 0.3
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    alpha = 0.4,
    size = 1.1
  ) +
  facet_wrap(~study_species) +
  labs(x = "Trial day", y = "latency to swim (s)", colour = "Condition", fill = "Condition") +
 
  theme_minimal(base_size = 13) +
  scale_y_log10()

```




CONTACT WITH SIDES or BOTTOM (called thigmotaxis here)
```{r}
ggplot(sub10, aes(x = trial_day, y =timespent_contact, colour = condition)) +
  geom_line(
    aes(group = interaction(individual_id, condition)),
    alpha = 0.12,
    position = position_dodge(width = 0.8)
  ) +
  geom_violin(
    aes(fill = condition),
    position = position_dodge(width = 0.8),
    alpha = 0.2,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    alpha = 0.3
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    alpha = 0.4,
    size = 1.1
  ) +
  facet_wrap(~study_species) +
  labs(x = "Trial day", y = "thigmotaxis (s)", colour = "Condition", fill = "Condition", title = "Thigmotaxis, first 10mins") +
  theme_minimal(base_size = 13)
```
 Jo's data not available yet. Possible to use overall and average per minute?



STARTLE

```{r}

ggplot(sub10, aes(x = trial_day, y =nb_dart_startle , colour = condition)) +
  geom_line(
    aes(group = interaction(individual_id, condition)),
    alpha = 0.12,
    position = position_dodge(width = 0.8)
  ) +
  geom_violin(
    aes(fill = condition),
    position = position_dodge(width = 0.8),
    alpha = 0.2,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    position = position_dodge(width = 0.8),
    alpha = 0.3
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    alpha = 0.4,
    size = 1.1
  ) +
  facet_wrap(~study_species) +
  labs(x = "Trial day", y = "number startle/darts", colour = "Condition", fill = "Condition", title = "startle/darts") +
  theme_minimal(base_size = 13)
  
```
Not relevant for most species. Day 1 should be same for Danio - fish used in 30 min condition not bothered. 
Experimental issue: Fish kept in communal lab, experiments done in separate temporal blocks - 10 min trials happened at very busy time! 30min trials: 2 weeks later, much calmer.



## Novel Object Tests

### Latency approach
```{r}

novO_data <- data %>%
  dplyr::filter(trial_day == "4")


ggplot(novO_data, aes(x = condition, y = latency_approach_no, colour = condition)) +
  geom_violin(fill = NA, trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.12), alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~study_species) +
  labs(
    x = "Acclimation condition",
    y = "Latency to approach novel object (log scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

```
Sparus/Maria: day 4 so far only available for 10min acclimation


### Latency touch
```{r}

ggplot(novO_data, aes(x = condition, y = latency_touch_no, colour = condition)) +
  geom_violin(fill = NA, trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.12), alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~study_species) +
  labs(
    x = "Acclimation condition",
    y = "Latency to touch novel object (log scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

```
Neoplamprologus: NO touch in 10min condition, only 2 in 30

### Distance

Only use as covariate  - that's fish release distance from novel object, not behaviour
```{r}

ggplot(novO_data, aes(x = condition, y = distance_no, colour = condition)) +
  geom_violin(fill = NA, trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.12), alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~study_species) +
  labs(
    x = "Acclimation condition",
    y = "distance novel object"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
  
```

